Historical Soil Organic Carbon Budget

### 2020-06-18 SOCbudgetPaper/introduction.Rmd
Update introduction.Rmd

The depletion of Soil organic carbon (SOC) due agricultural management has been a major source of carbon emissions, but also offers the potential to turn into a large additional carbon sink and thus mitigate climate change. Whereas the extent of the sink potential by SOC enhancement is highly debated (add refs), it is largely agreed upon that the SOC pool itself is the biggest terrestrial carbon pool, exceeding the atmospheric and even more the biospheric carbon pool multiple times. Even small changes in SOC drivers might lead to substantial shifts in earth carbon cycled and influence the atmospheric CO2 concentration (ref. permafrost melting). The specific amount of carbon stored in the soil is uncertain though and estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Bathjes, 1996).

This is critically important since managing agricultural soil might be one of the few setscrew for humans to _naturally_ combat climate change (ref. to natural climate solutions). Other _natural_ options, which might be summarized by restoring potential natural stocks via renaturation of managed land, seem to be often limited due to increased pressure on the land systems for feed, food and energy demand of a growing and developing population on earth.

Mapping the worlds SOC has been researched intensivly over the last decades and led to an increased quality of SOC maps as well as a even better understanding of factors driving magnitude, distribution and dynamics of SOC pools. It is undisputed that natural properties like climatic, biophysical and landscape characteristics play the most important role in this regard. Nevertheless,
these factors can not be easily altered by human intervention, leaving land cover, land use and land management changes the most important factors for current SOC dynamics. <!--to heroic find another word-->.

Recent studies identified the anthropogenic SOC debt at around 116 GtC (Sanderman et al.), which compares to older estimates of around 60-130 GtC <!--look up --> (Lal, 2006). Other studies have focused more closely on spatial disaggregated SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemcial processes within SOC dynamics  (<!-- find microbial study -->). Despite their importance in their field, these studies lack a detailed consideration of land management.

Also small-scale models (ref. Daycent, RothC, Ecosse, C-Tool) are able to represent well SOC dynamics on field-scale driven by detailed management data on yield levels, fertilizer inputs and various other on farming activities. Due to the lack of comprehensive global management data to feed these input demanding models, it seems still very complex to scale them up to a global extent.

Our study therefore estimates the change of global SOC stocks under specific consideration of agricultural mangement, including the impacts of different yield levels, residue recycling, manure amendments, irrigation and tillage. Our approach combines an spatial-explicit estimate of agricultural management data on global level with a robust but light weighted SOC model to estimate SOC stocks and stock change factors as well as organic carbon flow dynamics within the agricultural system.

>>>>

Soil organic carbon (SOC) is the biggest terrestrial carbon pool, exceeding the atmospheric and even more the biospheric carbon pool multiple times. Estimates of this carbon pool are still connected to large uncertainties, with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Bathjes, 1996). Even small changes in SOC can therefore lead to substantial shifts in earth carbon cycled and influence the atmospheric CO2 concentration (ref. permafrost melting).  Mapping the worlds SOC has been investigated intensivly over the last decades and led to an increased quality of SOC maps as well as a even better understanding of factors driving magnitude, distribution and dynamics of SOC pools. Natural properties like climatic, biophysical and landscape characteristics clearly play the most important role in this regard.

Human intervention, including land cover change and land management, has however added a further driver to SOC change, which alters terrestrial carbon pools in much shorter time scales and is likely the dominant driver of SOC change today. These  The depletion of Soil organic carbon (SOC) due to agricultural management has been a major source of carbon emissions. Recent studies identified the anthropogenic SOC debt at around 116 GtC (Sanderman et al.), which compares to older estimates of around 60-130 GtC <!--look up --> (Lal, 2006). Other studies have focused more closely on spatial disaggregated SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemcial processes within SOC dynamics  (<!-- find microbial study -->). Despite providing a rough estimate of the order of magnitude of change, these studies lack a detailed consideration of land management.

Field-scale models (ref. Daycent, RothC, Ecosse, C-Tool) are able to capture these land mangement impacts by using detailed information on yield levels, fertilizer inputs and various other on farming activities. Yet, due to the lack of comprehensive global management data to feed these input demanding models, it seems still very complex to scale them up to a global extent.

Our study combines an spatial-explicit estimate of agricultural management data on global level with a robust but light weighted SOC model to estimate SOC stocks and stock change factors as well as organic carbon flow dynamics within the agricultural system. We thereby consider change in SOC caused by historical land cover change as well as of different agricultural mangement practices, including residue recycling, manure amendments, irrigation and tillage. We thereby provide the first global, spatial explicit SOC maps that consider agricultural management.
